Assessing Power of the California Microhap Baseline for GSI (and calculating Fst)

### 2024-05-14 002-GSI_and_Fst.Rmd
commit before deleting a bunch

USING_SNAKEMAKE <- FALSE
NEG_USING_SMK <- TRUE
if(exists("snakemake")) {
  USING_SNAKEMAKE <- TRUE
  NEG_USING_SMK <- FALSE
}
```

```{r rmd_snakemake_hacks, include=FALSE, eval=USING_SNAKEMAKE}
# currently, this block does not get evaluated, while developing,
# but later when we snakemake-ize it all, we will evaluate it.
if(USING_SNAKEMAKE) {
  input_list <- snakemake@input
  output_list <- snakemake@output
}
```

@@

```{r input_bypass, include=NEG_USING_SMK, eval=NEG_USING_SMK}
# inputs:
input_list <- list(
  #final_baseline = "data/cali-chinook-baseline.rds",
  final_baseline = "data/subRoSA_baseline_rubias_FEB2024_w_mixture_revised.csv",
  lfar_genos = "data/metagenos_hap2col_lfar_all_samps.csv",
  pop_labels = "inputs/reference-collection-names.csv",
  map_notations = "inputs/map-notations.tsv"
)
# outputs:
output_list <- list(
  assig_table = "results/GSI_and_Fst/assignment-table-full-baseline.pdf",
  fst_table = "results/GSI_and_Fst/full-baseline-but-no-lfar-Fst-table.pdf",
  assig_table_lfar_all = "results/GSI_and_Fst/assignment-table-full-baseline-plus-lfar.pdf",
  tex_table = "tex/inputs/samples-table.tex",
  gsi_fst_fig = "results/GSI_and_Fst/gsi_and_fst_fig.pdf",
  gsi_fst_fig_crop = "results/GSI_and_Fst/gsi_and_fst_fig-crop.pdf",
  gsi_fst_fig_tex = "tex/images/gsi_and_fst_fig-crop.pdf"
)
# you can create the necessary output directories like this:
dump <- lapply(output_list, function(x)
  dir.create(dirname(x), recursive = TRUE, showWarnings = FALSE)
)
```

@@

# get the full baseline with the new markers, etc.  We will toss out some
# non-winter Sacto fish of unknown provenance
full_base <- read_csv(input_list$final_baseline) %>%
  filter(collection != "SacNonW")

@@

# finally, get the order of collections and repunits we want
collection_order <- unique(pop_labels$collection)
```

@@

@@

In this round, we go directly to assigning to repunit, and then to the max-likelihood
collection within that repunit.

@@

```{r}
top_ass <- full_sa %>%
  group_by(indiv, collection, repunit, inferred_repunit) %>%
  mutate(repu_sclike = sum(scaled_likelihood)) %>%
  group_by(indiv) %>%
  filter(repu_sclike == max(repu_sclike)) %>%
  arrange(indiv, desc(scaled_likelihood)) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(coll_f = factor(collection, levels = collection_order))
```

@@

And I want to look at the observed distribution of these scores in for each
comparison, colored by repunit.
```{r}
ggplot(focal_comparisons, aes(x = )
```

>>>>

@@

```{r}
if(exists("snakemake")) {
  input_list <- snakemake@input
  output_list <- snakemake@output
} else {
  input_list <- list(
    #final_baseline = "data/cali-chinook-baseline.rds",
    final_baseline = "data/subRoSA_baseline_rubias_FEB2024_w_mixture_revised.csv",
    final_baseline_apr = "data/subRoSA_baseline_rubias_APR2024_w_mixture_revised_rosa_added.csv",
    lfar_genos = "data/metagenos_hap2col_lfar_all_samps.csv",
    pop_labels = "inputs/reference-collection-names.csv",
    map_notations = "inputs/map-notations.tsv"
  )
  # outputs:
  output_list <- list(
    assig_table = "results/GSI_and_Fst/assignment-table-full-baseline.pdf",
    fst_table = "results/GSI_and_Fst/full-baseline-but-no-lfar-Fst-table.pdf",
    assig_table_lfar_all = "results/GSI_and_Fst/assignment-table-full-baseline-plus-lfar.pdf",
    tex_table = "tex/inputs/samples-table.tex",
    gsi_fst_fig = "results/GSI_and_Fst/gsi_and_fst_fig.pdf",
    gsi_fst_fig_crop = "results/GSI_and_Fst/gsi_and_fst_fig-crop.pdf",
    gsi_fst_fig_tex = "tex/images/gsi_and_fst_fig-crop.pdf"
  )
}

# you can create the necessary output directories like this:
dump <- lapply(output_list, function(x)
  dir.create(dirname(x), recursive = TRUE, showWarnings = FALSE)
)
```

@@

# get the full baseline with the new markers, etc.  We will toss out some
# non-winter Sacto fish of unknown provenance
###full_base <- read_csv(input_list$final_baseline) %>%
###  filter(collection != "SacNonW")

full_base_apr <- read_csv(input_list$final_baseline_apr) %>%
  filter(collection != "SacNonW")

# remove the rosa hapstr from it
full_base <- full_base_apr %>%
  select(-hapstr)

# capture that rosa hapstr with some meta data from future use
rosa_hapstrs <- full_base_apr %>%
  select(indiv:hapstr)

@@

# finally, get the order of collections and repunits we want
collection_order <- unique(pop_labels$collection)

@@

# Include the LFAR in there in a version

# Process the RoSA hapstrs to include at some point

We are not going to delve on the different between E1 and E2 here.  That is for the
subRoSA paper.  But we want to get the genotypes simply as EE, EL, LL as done in the
science paper.

Let us have a quick look at the hapstrs:
```{r}
rosa_hapstrs %>%
  count(hapstr) %>%
  arrange(desc(n))
```

We will explode those into a vector of characters at positions
`c(1, 3, 4, 6:10)`, thus dropping the E2-specific variants as
well as the Tasha2 SNP that effectively tags the duplication, leaving
us with the 8 core RoSA SNPs.
```{r}
H8 <- rosa_hapstrs %>%
  select(indiv, hapstr) %>%
  mutate(
    hvec = map(hapstr, function(x) {y <- strsplit(x, "")[[1]]; y[c(1, 4, 5, 7:11)]}),
    hapstr8 = map_chr(hvec, function(x) paste(x, collapse = "", sep = ""))
  )

# Let's count up those hapstr8's
H8 %>%
  count(hapstr8) %>%
  arrange(desc(n))
```

From that it is clear that there have been some recombinations in here, but we will
just go ahead and call RoSA genotype by the majority of the non-missing ones. 
```{r}
H8_strip <- H8 %>%
  mutate(
    stripped = map(hvec, function(x) x[!(x %in% c("?", "N"))]),
    sglom = map_chr(stripped, function(x) paste(x, collapse = "", sep = ""))
    )

H8_strip %>%
  count(sglom) %>%
  arrange(desc(n))

OK, we will call something and EL if it has two or more Hs (the idea being that a single
H could be a genotyping error). Note that there are no haplotypes mixing L's and E's together.
Here is a function that will determine the RoSA genotype
from a vector of H, L, and E parts. 
```{r}
call_rosa <- function(x) {
  # things that were all ?'s get NA
  if(length(x) == 0) {
    return(NA_character_)
  }

  nE <- sum(x == "E")
  nL <- sum(x == "L")
  nH <- sum(x == "H")

  if(nH >= 2) { 
    return("EL")
  } else if(nL > 0 && nE == 0) {
    return("LL")
  } else if(nE > 0 && nL == 0) {
    return("EE")
  } else {
    return("Weird")
  }

}

H8_genos <- H8_strip %>%
  mutate(rosa_geno = map_chr(stripped, call_rosa))
```

Count these up so we can see what we called LL, EL, and EE:
```{r}
H8_genos %>%
  count(sglom, rosa_geno) %>%
  arrange(desc(n))
```

Now we will slim it down for future use:
```{r}
rosa_genos <- H8_genos %>%
  select(indiv, rosa_geno)
```

@@

We go directly to assigning to repunit, and then to the max-likelihood
collection within that repunit.  We include the lfar in here.

@@

```{r}
source("R/get_top_assignments.R")

top_ass <- get_top_assignments(full_sa, collection_order)
```

Plot that initial table:
```{r}
source("R/plot_self_assigment_table.R")

no_lfar_ass_table <- plot_self_assignment_table(top_ass, pop_labels, collection_order)

no_lfar_ass_table_80 <- top_ass %>% 
  filter(scaled_likelihood > 0.8) %>%
  plot_self_assignment_table(pop_labels, collection_order)

no_lfar_ass_table_Rosa <- top_ass %>%
  left_join(rosa_genos, by = join_by(indiv)) %>%
  plot_self_assignment_table(pop_labels, collection_order)
```

And then do the same for the data with the LFAR:
```{r}
lfar_sa <- self_assign(base_with_lfar, gen_start_col = 5)
lfar_top_ass <- get_top_assignments(lfar_sa, collection_order)
lfar_ass_table <- plot_self_assignment_table(lfar_top_ass, pop_labels, collection_order)

lfar_ass_table_80 <- lfar_top_ass %>% 
  filter(scaled_likelihood > 0.8) %>%
  plot_self_assignment_table(pop_labels, collection_order)

lfar_ass_table_Rosa <- lfar_top_ass %>%
  left_join(rosa_genos, by = join_by(indiv)) %>%
  plot_self_assignment_table(pop_labels, collection_order)
```

@@

And I want to look at the observed distribution of these scores in for each
comparison, colored by repunit.
```{r}
ggplot(focal_comparisons, aes(x = ))
```
